On the spatial aggregation of ecosystem condition indicators

### 2025-09-22 manuscript/manuscript.qmd
Hannos new example

<!-- Some indicators are possible to normalise at the measurement scale, 
i.e. the finest resolution of the raw data.
Others, however, need to be (pre-) aggregated, 
for example if the variable is designed to depict a density or a proportion that refers to a spatial unit. 
Generally, variables that describe "the share of X in Y" 
(e.g. the proportion of one type of "surface cover" within a larger area) 
are particularly prone to issues regarding scale dependence of the reference levels.
If you look at such a variable for a large area, then you can extract some meaningful and precise etimate about the proportion of X in Y, 
but for increasingly smaller areas the variable become more binary, until you have either only Y or no Y present.
At this zoom level the reference levels is unlikely to make much sense. 
This stresses the need to define ecosystem assets that are appropriate in size. -->

@@

::: callout-note
**Example 3 -- Phytoplankton trophic index (PTI)**\

*Variable*: Mean score of algal species present, based on a set of indicator species scored for phosphorus requirements/tolerance.
The variable is recorded in water bodies (lakes of 0.5 km^2^ or more).

*Reference levels*: X~100~ = median variable value for water bodies in reference condition.
X~0~, X~20~, X~40~, X~60~ and X~80~ = intercalibrated threshold values, based on dose–response curves.\

```{r read}
#| include: false
#| cache: false
dat <- readxl::read_xlsx(here::here("data/PTI_data.xlsx")) |>
  mutate(X = X |> str_replace_all("%", "\\\\%"))
cn <- colnames(dat) |>
    str_replace_all("2", "") |>
    str_replace_all("X", "") |>
    str_replace_all("RL", "X\\\\textsubscript{100}")
```

```{r tbl-pti}
#| echo: false
#| tbl-cap: "Dummy data for the PTI example. value = value for a hypothetical variable. agg. = spatial aggregation (arithmetic mean); trunc. truncation; stat. = spatial statistics. The transformation step is left out for simplicity. Pathways refer to figure 3."
#| cache: false
kbl(dat, format = "latex",
    booktabs = TRUE,
    col.names = cn,
    escape = F
    ) |>
  kable_styling() %>%
  add_header_above(c(
    " " = 3, 
    "Pathway 5" = 4,
     "Pathway 6" = 3)) |>
  collapse_rows(
    columns = c(9,10),
    latex_hline = "major"
  )
```

The Water Framework Directive (WFD) indicators uses aggregation pathway 5 (@fig-pathways), with truncation, scaling, and transformation, in that order (see worked example in @tbl-pti).
The value obtained after the first two steps is called an EQR (ecological quality ratio), and the value obtained after the third step nEQR (normalised EQR).
Spatial aggregation, the way that we use the term here, is not done for the WFD.
Instead, the proportion of water bodies in each condition class is the aggregated reporting metric.
Therefore aggregation displacement is not an issue.
When WFD data is put into use in other contexts, however, aggregation becomes important.

\
Because of how X~100~ is defined based on the median value across reference lakes, overshooting values (\> X~100~) should be preserved in the spatial aggregation.
Otherwise we get a negative displacement, making it practically impossible to reach an indicator value of 1.
This means that neither EQR values nor nEQR values can be uncritically aggregated spatially to be used in a SEEA EA compliant in ECA.
One solution, to enable the use of WFD indicators in ECAs, would be to change to pathway 6, with (1) scaling, (2) transforming, (3) aggregating and (4) truncating the values.
In our hypothetical example (@tbl-pti) this gives an aggregated value of 0.85, whereas the mean value for the nEQRs would be 0.75 (i.e. negatively displaced).
:::

<!-- SEEA EA uses the term aggregation commutativity for the property that the order of vertical and horizontal aggregation is irrelevant to the end result.
A similar aggregation commutativity can be envisioned for step-wise aggregation of a single indicator, where it does not matter if the original data is aggregated directly from the original measurement scale to the Ecosystem Accounting Area (EAA) level, or via some intermittent scale, such as ecosystem assets or larger regions.
This is possible when aggregating metrics using area weighting, and when one is aware of the need for post-stratification when using stratified sampling. -->

<!-- Secondly, SEEA EA mentions several arithmetic or parametric aggregation functions, such as different means, median, min or max values.
It does not mention the possibility of bootstrap resampling [@jakobsson2021; @kolstad_making_2024].
This method allows for using probability distributions to represent indicator values and uncertainties, and to use weighted resampling to create new joint probability distributions for the spatially aggregated indicator. -->

@@

```{r fig-pathways}
#| fig-cap: |
#|   Flowchart showing pathways for the spatial aggregation of variables in ecosystem condition accounting. Normalisation refers to the steps that make a variable conform to the definition of an indicator in SEEA EA, which may include some, but not necesarily all the steps scale, truncate and transform.
#| out-width: 400
#| echo: false
knitr::include_graphics("../img/aggregationPathways.png")
```

>>>>

```{=html}
<!-- Some indicators are possible to normalise at the measurement scale, 
i.e. the finest resolution of the raw data.
Others, however, need to be (pre-) aggregated, 
for example if the variable is designed to depict a density or a proportion that refers to a spatial unit. 
Generally, variables that describe "the share of X in Y" 
(e.g. the proportion of one type of "surface cover" within a larger area) 
are particularly prone to issues regarding scale dependence of the reference levels.
If you look at such a variable for a large area, then you can extract some meaningful and precise etimate about the proportion of X in Y, 
but for increasingly smaller areas the variable become more binary, until you have either only Y or no Y present.
At this zoom level the reference levels is unlikely to make much sense. 
This stresses the need to define ecosystem assets that are appropriate in size. -->
```

@@

::: callout-note
**Example 3 -- Water Framework Directive, indicator ASPT**\

*Variable*: ASPT (*average score per taxon*) is the mean eutrophication tolerance score of invertebrate species present in a river water body.\

*Reference levels*: X~100~ = median variable value recorded in reference water bodies.
X~80~ = lower boundary of variable values in reference water bodies.
X~0~ = natural zero (no species present).
X~20~, X~40~, and X~60~ = intercalibrated threshold values, based on dose–response curves.\

The Water Framework Directive (WFD) indicators use aggregation pathway 5 (@fig-pathways), 
with truncation, scaling, and transformation, in that order.
The value obtained after the first two steps is called an EQR (ecological quality ratio), 
and the value obtained after the third step is called nEQR 
(normalised EQR, and true indicator values *sensu* SEEA EA).
Spatial aggregation, the way that we use the term here, is not done for the WFD.
Instead, the aggregated reporting metric is the proportion of water bodies in each of the five condition classes 
("bad", "poor", "moderate", "good", "high"), 
separated by the non-terminal reference levels. \

When WFD data is put into use in other contexts, however, aggregation becomes important, 
and it is crucial that aggregation is done after the (scaling and) transformation step, 
because the transformation (using a stepwise-linear function) adds ecological meaning to the data.
As the example in @fig-aspt shows, 
the ecologically meaningful interpretation is that the EAA has "poor" condition at time 1, improving to "moderate" condition at time 2.
When aggregation is erroneously done before transformation, however, 
the same measurements suggest that condition is "bad" and deteriorating.
This illustrates that an ecosystem condition indicator account cannot be derived directly from an ecosystem condition variable account in conjunction with reference levels, 
as suggested by SEEA EA (@unitednations2024: 5.76 and table 5.3).
The reason in the case of ASPT is that the transformation function is non-linear,
and aggregation displacement therefore occurs in the spatial aggregation of variables and indicators.
Reference levels are the same throughout the entire EAA, and variation in reference levels is thus not part of the problem here. \

![Different aggregation pathways can result in different indicator values, exemplified by the ASPT indicator from the WFD. Two ecosystem assets (Rivers A and B) are measured at two points in time. Regular font numbers are EQR values; boldface numbers are nEQR values (indicator values *sensu* SEEA EA). When EQR values are first aggregated and then transformed ("transf."), ASPT incorrectly indicates a deteriorating and "bad" condition. When EQR values are first transformed and then aggregated, ASPT correctly indicates an improvement from "poor" to "moderate" condition. ](images/ASPT.jpg){#fig-aspt}

Another issue that can be illustrated with ASPT (or other WFD indicators) is the effect of truncation.
Aggregation pathway 5 uses truncation as the first step. 
When WFD data is used in ecosystem condition accounting, however, 
truncation needs to be the final step, 
applied only after aggregation. 
This is because X~100~ is defined as the median of variable values measured in reference water bodies. 
In other words, half of the values measured in reference water bodies is expected to be above X~100~. 
When truncation is done as the first step, 
the average indicator value of reference water bodies will be around 0.95, 
i.e. it would be prevented from actually reaching 1.0. 
Overshooting values therefore need to be preserved until after aggregation (spatial compensation),
and only after aggregation should overshooting values be truncated to 1.0. \

Likewise, in the case where model-based extrapolation of a WFD indicator is used, 
it is important to postpone truncation until after the modelling and extrapolation. 
Otherwise the distribution of the extrapolated values may be severely distorted, 
at least if many of the original measurements are in or close to the X~100~ value. 
:::

```{=html}
<!-- SEEA EA uses the term aggregation commutativity for the property that the order of vertical and horizontal aggregation is irrelevant to the end result.
A similar aggregation commutativity can be envisioned for step-wise aggregation of a single indicator, where it does not matter if the original data is aggregated directly from the original measurement scale to the Ecosystem Accounting Area (EAA) level, or via some intermittent scale, such as ecosystem assets or larger regions.
This is possible when aggregating metrics using area weighting, and when one is aware of the need for post-stratification when using stratified sampling. -->
```

```{=html}
<!-- Secondly, SEEA EA mentions several arithmetic or parametric aggregation functions, such as different means, median, min or max values.
It does not mention the possibility of bootstrap resampling [@jakobsson2021; @kolstad_making_2024].
This method allows for using probability distributions to represent indicator values and uncertainties, and to use weighted resampling to create new joint probability distributions for the spatially aggregated indicator. -->
```

@@

```{r fig-pathways}
#| fig-cap: |
#|   Flowchart showing pathways for the spatial aggregation of variables in ecosystem condition accounting. Normalisation refers to the steps that make a variable conform to the definition of an indicator in SEEA EA, which may include some, but not necessarily all, of the steps scale, truncate and transform.
#| out-width: 400
#| echo: false
knitr::include_graphics("../img/aggregationPathways.png")
```
